---
title: "Logistic Regression: Interpreting Results"
date: today
date-format: long
footer: "[IAFF 6501 Website](https://quant4ia.rocks/)"
logo: images/iaff6501-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(peacesciencer)

conflict_df <- create_stateyears(system = 'gw') |>
  filter(year %in% c(1946:1999)) |>
  add_ucdp_acd(type=c("intrastate"), only_wars = FALSE) |>
  add_democracy() |>
  add_creg_fractionalization() |>
  add_sdp_gdp() |>
  add_rugged_terrain()

conflict_model <- glm(ucdponset ~ wbgdppc2011est,
                      data = conflict_df,
                      family = "binomial")
```

# Interpreting Logistic Regression Results

## The Model Equation

<br>

$$\log\left(\frac{p}{1-p}\right) = -1.16-0.33\times \text{logGDPpc}$$

## What Are Log-Odds?

<br>

::: {.incremental}
- Logistic regression coefficients are on the **log-odds** scale
- The direction of the effect is easy to read (positive = more likely, negative = less likely)
- But the magnitude is hard to interpret directly
- We need to transform the coefficients into something more intuitive
:::

## Odds Ratios

<br>

- For a more intuitive interpretation, we can **exponentiate** the coefficients
- The exponentiated coefficient is the **odds ratio**
- For each one-unit increase in the predictor, the odds of the outcome are **multiplied** by the odds ratio

## Interpreting the Odds Ratio {.smaller}

<br>

$$\log\left(\frac{p}{1-p}\right) = -1.16-0.33\times \text{logGDPpc}$$

<br>

For each one unit increase in log GDP per capita, the odds of conflict onset are multiplied by approximately 0.718, assuming other variables are held constant.

<br>

This means that an increase in GDP per capita is associated with a **decrease** in the odds of conflict onset. The odds decrease by about 28.2% for each unit increase in log GDP per capita.

## Calculating Odds Ratios in R

<br>

```{r}
#| label: odds_ratios
#| echo: true

library(broom)

tidy(conflict_model, exponentiate = TRUE) |>
  select(term, estimate, p.value)
```

## Your Turn! {.smaller}

<br>

- Run a bivariate logistic regression with a different predictor (e.g., `v2x_polyarchy`, `ethfrac`)
- Use `tidy(..., exponentiate = TRUE)` to get the odds ratios
- Interpret the results:
  - Is the effect positive or negative?
  - How much do the odds of conflict onset change per unit increase?

```{r}
#| label: time1

library(countdown)

countdown(minutes = 10,
          id = "timer1",
          bottom = "5%",
          right = "10%",
          color_border = "#fff",
          color_text = "#fff",
          color_running_background = "#42affa",
          color_running_text = "black",
          color_finished_background = "#E5D19D",
          color_finished_text = "#00264A")
```

# Predicted Probabilities

## Calculating Predicted Probabilities {.scrollable}

<br>

Probability of conflict onset for a country with log GDP per capita of 9 (about $8,000):

$$\log\left(\frac{p}{1-p}\right) = -1.16-0.33\times 9 = -4.13$$

$$\frac{p}{1-p} = \exp(-4.13) = 0.016$$

$$p = \frac{0.016}{1 + 0.016} = 0.0158$$

## Using `marginaleffects`

```{r}
#| label: marginal_effects
#| echo: true
#| eval: false

library(marginaleffects)

# select some countries for a given year
selected_countries <- conflict_df |>
  filter(
    gw_name %in% c("United States of America", "Venezuela", "Rwanda"),
    year == 1999)

# calculate predicted probabilities
marg_effects <- predictions(conflict_model, newdata = selected_countries)

tidy(marg_effects) |>
  select(estimate, p.value, conf.low, conf.high, gw_name)
```

## Using `marginaleffects`

<br>

```{r}
#| label: marginal_effects2

library(marginaleffects)

prediction_data <- data.frame(wbgdppc2011est = c(6, 7, 8, 9, 10))
pred_results <- predictions(conflict_model, newdata = prediction_data)

tidy(pred_results) |>
  select(wbgdppc2011est, estimate, conf.low, conf.high) |>
  mutate(GDP_approx = paste0("~$", round(exp(wbgdppc2011est))))
```

## Your Turn!

<br>

- Select your favorite three countries and a recent year
- Calculate the predicted probability of conflict onset using the `marginaleffects` package
- If you have time, try to verify one calculation by hand

```{r}
#| label: time2

countdown(minutes = 10,
          id = "timer2",
          bottom = "10%",
          right = "10%")
```
