---
title: "Logistic Regression: Motivation and Model"
date: today
date-format: long
footer: "[IAFF 6501 Website](https://quant4ia.rocks/)"
logo: images/iaff6501-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---


```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(peacesciencer)
```

# Binary Outcomes

## Link to Code

<br>

Please find the code [here](https://www.dropbox.com/scl/fo/g4tdpdwcij78nyydvr4bp/h?rlkey=hctskhle6222csnwre5ir9gz3&dl=0)

## Binary Outcomes

<br>

- So far we have looked at continuous or numerical outcomes (response variables)
- We are often also interested in outcome variables that are binary (Yes/No, or 1/0)
    - Did violence happen, or not?
    - Classification: is this email spam?

## Example: Conflict Onset

<br>

::: {.incremental}
- Did a civil war begin in a given country in a given year? (yes/no)
- Predictors: wealth, democracy, terrain, ethnic diversity, etc. 
- Seminal work by Fearon and Laitin (2003) 
- We can use logistic regression to model this binary outcome
:::

## Modeling

<br>

::: {.incremental}
- We can treat each outcome (conflict onset) as successes and failures arising from separate Bernoulli trials
- **Bernoulli trial**: a random experiment with exactly two possible outcomes, "success" and "failure", in which the probability of success is the same every time the experiment is conducted
- Success is usually coded as 1, failure as 0
- So ironically, conflict onset is a "success" in this context
:::

## Modeling

<br>

Each Bernoulli trial can have a separate probability of success

<br>

$$ y_i ∼ Bern(p) $$

## Modeling

<br>

- We can then use the predictor variables to model that probability of success, $p_i$
- We can't really use a linear model for $p_i$ (since $p_i$ must be between 0 and 1) but we can transform the linear model to have the appropriate range


## Generalized Linear Models

<br>

- This is a very general way of addressing many problems in regression and the resulting models are called **generalized linear models (GLMs)**
- Logistic regression is a very common example

## GLMs

<br>

All GLMs have the following three characteristics:

::: {.incremental}
- A probability distribution describing a generative model for the outcome variable
- A linear model:
$$\eta = \beta_0 + \beta_1 X_1 + \cdots + \beta_k X_k$$
- A link function that relates the linear model to the parameter of the outcome distribution
:::
  
## Logistic Regression

<br>

- Logistic regression is a GLM used to model a binary categorical outcome (0 or 1)
- In logistic regression, the link function that connects $\eta_i$ to $p_i$ is the **logit function**
- **Logit function:** For $0\le p \le 1$

$$logit(p) = \log\left(\frac{p}{1-p}\right)$$

## Logit Function


```{r}
#| label: logit_function

d <- tibble(p = seq(0.001, 0.999, length.out = 1000)) %>%
  mutate(logit_p = log(p/(1-p)))

ggplot(d, aes(x = p, y = logit_p)) + 
  geom_line() + 
  xlim(0,1) + 
  ylab("logit(p)") +
  labs(title = "logit(p) vs. p")
```


## Logistic Regression Model

<br>

::: {.incremental}
- $y_i \sim \text{Bern}(p_i)$
- $\eta_i = \beta_0+ \beta_1 x_{1,i} + \cdots + \beta_n x_{n,i}$
- $\text{logit}(p_i) = \eta_i$
:::

## Logistic Regression Model

<br>

- $\text{logit}(p_i) = \eta_i = \beta_0+ \beta_1 x_{1,i} + \cdots + \beta_n x_{n,i}$
- Now take inverse logit to get $p$

$$p_i = \frac{\exp(\beta_0+\beta_1 x_{1,i} + \cdots + \beta_k x_{k,i})}{1+\exp(\beta_0+\beta_1 x_{1,i} + \cdots + \beta_k x_{k,i})}$$

# Analyzing Conflict Onset

## The `peacesciencer` Package

- The `peacesciencer` package provides a number of datasets and functions for analyzing conflict and peace
- Provides data from a number of important datasets in the field of conflict studies, e.g.
  - Correlates of War (CoW) project
  - Uppsala Conflict Data Program (UCDP)
  - Militarized Interstate Dispute (MID) dataset
- Provides functions for analyzing conflict and adding control variables to the dataset

## Using the `peacesciencer` Package

<br>

```{r}
#| label: use_peacesciencer
#| echo: true

library(peacesciencer)

conflict_df <- create_stateyears(system = 'gw') |>
  filter(year %in% c(1946:1999)) |>
  add_ucdp_acd(type=c("intrastate"), only_wars = FALSE) |>
  add_democracy() |>
  add_creg_fractionalization() |>
  add_sdp_gdp() |>
  add_rugged_terrain()

glimpse(conflict_df)
```

## Running a Logistic Regression

<br>

- Implementation is not very different from a linear model
- Instead of `lm()`, we use `glm()` with `family = "binomial"`
- The `family` argument specifies the logit link function automatically

## Bivariate Logistic Regression

<br>

```{r}
#| label: bivariate_model
#| echo: true

conflict_model <- glm(ucdponset ~ wbgdppc2011est,
                      data = conflict_df,
                      family = "binomial")

summary(conflict_model)
```

## Your Turn!

<br>

- Run a bivariate logistic regression using `ucdponset` as the outcome variable
- Try a different predictor (e.g., `v2x_polyarchy`, `ethfrac`, `rugged`)
- Examine the model output with `summary()`
- What direction is the effect? Is it significant?

```{r}
#| label: time1

library(countdown)

countdown(minutes = 10,
          id = "timer1",
          bottom = "5%",
          right = "10%",
          color_border = "#fff",
          color_text = "#fff",
          color_running_background = "#42affa",
          color_running_text = "black",
          color_finished_background = "#E5D19D",
          color_finished_text = "#00264A")
```

