---
title: Outliers and Model Assumptions
date: today
date-format: long
footer: "[IAFF 6501 Website](https://quant4ia.rocks/)"
logo: images/iaff6501-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: false
  message: false
  warning: false
  freeze: auto
---

```{r}
#| label: setup

library(tidyverse)
library(vdemlite)
library(broom)

model_data <- fetchdem(
  indicators = c("v2x_libdem", "e_gdppc", "v2cacamps"),
  start_year = 2019,
  end_year = 2019
) |>
  rename(
    country = country_name,
    lib_dem = v2x_libdem,
    wealth = e_gdppc,
    polarization = v2cacamps
  ) |>
  filter(!is.na(lib_dem), !is.na(wealth))
```

# Dealing with Outliers

## What Are Outliers?

<br>

- Observations that don't fit the general pattern of the data
- Not always errors — often **substantively interesting cases**
- In our GDP-democracy example: oil-rich authoritarian states

## Three Types of Outliers {.smaller}

<br>

- **Leverage points**: extreme values on the x-axis
  - Unusual predictor values; *potential* to distort the model
- **Influential points**: significantly change the regression line when removed
  - Combine extreme predictors *and* large residuals
- **Residual outliers**: points far from the regression line
  - Poorly fit by the model, regardless of their x-value

## Revisiting Our GDP-Democracy Data

```{r}
#| label: scatter-plot

library(ggrepel)

model_data_labeled <- model_data |>
  mutate(
    label = case_when(
      wealth > 50 | (wealth > 20 & lib_dem < 0.25) ~ country,
      TRUE ~ NA_character_
    )
  )

ggplot(model_data_labeled, aes(x = wealth, y = lib_dem)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(
    aes(label = label),
    size = 3,
    max.overlaps = 20,
    seed = 42
  ) +
  scale_x_log10(labels = scales::label_dollar(suffix = "k")) +
  labs(x = "GDP per Capita (log scale)",
       y = "Liberal Democracy Score",
       title = "Wealth and Democracy, 2019") +
  theme_minimal()
```

## Identifying Outliers: IQR Method

```{r}
#| label: boxplots
#| echo: true
#| code-fold: true

library(patchwork)

p1 <- ggplot(model_data, aes(y = lib_dem)) +
  geom_boxplot() +
  labs(y = "Liberal Democracy Score", title = "Democracy") +
  theme_minimal()

p2 <- ggplot(model_data, aes(y = wealth)) +
  geom_boxplot() +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(y = "GDP per Capita", title = "GDP (Raw)") +
  theme_minimal()

p1 | p2
```

## Flagging GDP Outliers {.smaller}

<br>

```{r}
#| label: flag-outliers
#| echo: true

model_data <- model_data |>
  mutate(gdp_outlier = wealth %in% boxplot.stats(wealth)$out)

model_data |>
  filter(gdp_outlier) |>
  arrange(desc(wealth)) |>
  select(country, wealth, lib_dem)
```

## Regression Diagnostics

<br>

The `broom::augment()` function adds diagnostic columns to our data:

| Column | What it measures |
|--------|-----------------|
| `.hat` | **Leverage** — how extreme the predictor values are |
| `.std.resid` | **Standardized residuals** — how far from the line |
| `.cooksd` | **Cook's distance** — overall influence on the model |

## Flagging Influential Points {.smaller}

<br>

```{r}
#| label: diagnostics
#| echo: true

democracy_model <- lm(lib_dem ~ log(wealth), data = model_data)

model_diagnostics <- augment(democracy_model, data = model_data) |>
  mutate(
    high_leverage  = .hat > 2 * mean(.hat),
    high_residual  = abs(.std.resid) > 2,
    high_influence = .cooksd > 4 / nrow(model_data)
  )

model_diagnostics |>
  filter(high_leverage | high_residual | high_influence) |>
  select(country, wealth, lib_dem, high_leverage, high_residual, high_influence)
```

## Your Turn! {.smaller}

<br>

- Run the model using `polarization` as the predictor instead of `wealth`
- Use `augment()` to identify influential observations in this new model
- What countries stand out? Are they the same as in the GDP model?

```{r}
#| label: timer1

library(countdown)
countdown(minutes = 10, id = "timer1", right = "10%", bottom = "5%")
```

# Outlier Strategies

## Four Main Strategies

<br>

::: {.incremental}
1. **Leave them in** — outliers may be substantively important
2. **Remove them** — check how much they change your results
3. **Transform the data** — e.g., log transformation compresses extreme values
4. **Winsorize** — cap extreme values at a percentile rather than removing them
:::

## Strategy 1: Removing Outliers {.smaller}

<br>

Compare models with and without influential points:

```{r}
#| label: remove-outliers
#| echo: true

# Original model
model_full <- lm(lib_dem ~ log(wealth), data = model_data)

# Model without influential points
model_data_clean <- model_diagnostics |>
  filter(!high_leverage & !high_residual & !high_influence)

model_clean <- lm(lib_dem ~ log(wealth), data = model_data_clean)

# Compare coefficients
tibble(
  model = c("Full", "No outliers"),
  intercept = c(coef(model_full)[1], coef(model_clean)[1]),
  slope = c(coef(model_full)[2], coef(model_clean)[2]),
  r_squared = c(glance(model_full)$r.squared, glance(model_clean)$r.squared)
)
```

## Strategy 2: Log Transformation

<br>

```{r}
#| label: log-comparison

model_no_log  <- lm(lib_dem ~ wealth,       data = model_data)
model_with_log <- lm(lib_dem ~ log(wealth), data = model_data)

p1 <- ggplot(model_data, aes(x = wealth, y = lib_dem)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(title = "Raw GDP — outliers distort the line",
       x = "GDP per Capita", y = "Liberal Democracy") +
  theme_minimal()

p2 <- ggplot(model_data, aes(x = log(wealth), y = lib_dem)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Log GDP — much better fit",
       x = "Log GDP per Capita", y = "Liberal Democracy") +
  theme_minimal()

p1 | p2
```

## Strategy 3: Winsorizing {.smaller}

**Winsorizing** caps extreme values at a specified percentile instead of removing them:

```{r}
#| label: winsorize
#| echo: true
#| code-fold: true

library(datawizard)

model_data_win <- model_data |>
  mutate(wealth_win95 = winsorize(wealth, threshold = 0.05))

p1 <- ggplot(model_data, aes(x = wealth)) +
  geom_histogram(bins = 30, fill = "darkblue", alpha = 0.7) +
  scale_x_continuous(labels = scales::label_dollar(suffix = "k")) +
  labs(title = "Original", x = "GDP per Capita") +
  theme_minimal()

p2 <- ggplot(model_data_win, aes(x = wealth_win95)) +
  geom_histogram(bins = 30, fill = "darkred", alpha = 0.7) +
  scale_x_continuous(labels = scales::label_dollar(suffix = "k")) +
  labs(title = "Winsorized at 95th pct", x = "GDP per Capita") +
  theme_minimal()

p1 | p2
```

## Strategy 4: Robust Regression {.smaller}

<br>

`MASS::rlm()` is less sensitive to outliers than OLS:

```{r}
#| label: robust-regression
#| echo: true

library(MASS)

robust_model <- rlm(lib_dem ~ log(wealth), data = model_data)

summary(robust_model)
```

## How to Choose? {.smaller}

<br>

| Situation | Strategy |
|-----------|----------|
| Data entry error or wrong population | Remove |
| Naturally skewed variable | Log transform |
| Want to keep all obs., reduce tail influence | Winsorize |
| Outliers are legitimate but numerous | Robust regression |
| Outliers are theoretically interesting | Leave in; discuss |

<br>

::: {.callout-note}
Always report your decision and run a **sensitivity analysis** to show how results change.
:::

## Your Turn! {.smaller}

<br>

- Go back to the polarization model you built earlier
- Try removing the influential points and compare to the original model
- Now try winsorizing polarization at the 95th percentile
- Which approach changes the results more?

```{r}
#| label: timer2

countdown(minutes = 10, id = "timer2", right = "10%", bottom = "5%")
```

# Checking Model Assumptions (Optional)

## Why Check Assumptions?

<br>

::: {.incremental}
- Linear regression makes several assumptions about your data
- Violating them can lead to **biased coefficients** or **wrong standard errors**
- Checking assumptions helps you know *when to trust your results*
:::

## The LINE Conditions {.smaller}

<br>

| Letter | Assumption | What it means |
|--------|-----------|---------------|
| **L** | Linearity | Relationship between X and Y is linear |
| **I** | Independence | Observations are independent of each other |
| **N** | Normality | **Residuals** are approximately normally distributed |
| **E** | Equal variance | Residuals have constant variance (**homoscedasticity**) |

## Setup for Diagnostics

<br>

```{r}
#| label: setup-diagnostics
#| echo: true

library(vdemlite)

model_data2 <- fetchdem(
  indicators = c("v2x_libdem", "e_gdppc", "v2cacamps",
                 "v2x_gender", "v2x_corr", "e_regionpol_6C"),
  start_year = 2006, end_year = 2006
) |>
  rename(country = country_name, lib_dem = v2x_libdem,
         wealth = e_gdppc, polarization = v2cacamps,
         women_emp = v2x_gender, corruption = v2x_corr,
         region = e_regionpol_6C) |>
  mutate(region = factor(region,
    labels = c("Eastern Europe", "Latin America", "MENA",
               "SS Africa", "The West", "Asia & Pacific")))

democracy_model2 <- lm(lib_dem ~ log(wealth) + polarization +
                         corruption + women_emp + region,
                       data = model_data2)
```

## Exploratory Data Analysis First

```{r}
#| label: ggpairs

library(GGally)

model_data2 |>
  dplyr::select(lib_dem, wealth, polarization, corruption, women_emp) |>
  ggpairs(progress = FALSE)
```

## Checking Linearity: Residuals vs. Fitted

```{r}
#| label: resid-vs-fitted

diag_data <- model_data2 |>
  drop_na() |>
  mutate(
    fitted    = fitted(democracy_model2),
    residuals = residuals(democracy_model2)
  )

ggplot(diag_data, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(se = FALSE, color = "blue") +
  labs(x = "Fitted Values", y = "Residuals",
       title = "Residuals vs. Fitted — check for patterns") +
  theme_minimal()
```

## Checking Normality: Q-Q Plot

```{r}
#| label: qq-plot

qq_data <- diag_data |>
  arrange(residuals) |>
  mutate(theoretical = qnorm(ppoints(length(residuals))))

ggplot(qq_data, aes(x = theoretical, y = sort(residuals))) +
  geom_point() +
  geom_abline(slope = sd(diag_data$residuals),
              intercept = mean(diag_data$residuals),
              color = "red") +
  labs(title = "Normal Q-Q Plot — points should follow the line",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()
```

## Checking Equal Variance: Scale-Location

```{r}
#| label: scale-location

diag_data |>
  mutate(sqrt_abs_resid = sqrt(abs(residuals))) |>
  ggplot(aes(x = fitted, y = sqrt_abs_resid)) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE, color = "blue") +
  labs(x = "Fitted Values", y = "√|Residuals|",
       title = "Scale-Location — horizontal line = equal variance") +
  theme_minimal()
```

## Automated Diagnostics with `performance`

<br>

You can automate all the diagnostics we just discussed with a single line of code using the `performance` package:

```{r}
#| label: check-model-code
#| echo: true
#| eval: false

library(performance)

check_model(democracy_model2)
```

##

```{r}
#| label: check-model-plot
#| fig-height: 7
#| fig-width: 11
#| out-width: "100%"

library(performance)

check_model(democracy_model2)
```

## Fixing Violations: Transformations {.smaller}

<br>

If you detect assumption violations:

::: {.incremental}
- **Log transform** skewed predictors → compresses extremes, improves linearity
- **Polynomial terms** → capture curved relationships: `I(x^2)`
- **Robust standard errors** → correct inference under heteroscedasticity
:::

## Example: Addressing Non-Linearity

<br>

Adding a polynomial term for `women_emp` to capture a curved relationship:

```{r}
#| label: improved-model-code
#| echo: true
#| eval: false

democracy_model3 <- lm(
  lib_dem ~ log(wealth) + polarization + corruption +
    log(women_emp) + I(log(women_emp)^2) + region,
  data = model_data2
)

check_model(democracy_model3)
```

##

```{r}
#| label: improved-model-plot
#| fig-height: 7
#| fig-width: 11
#| out-width: "100%"

democracy_model3 <- lm(
  lib_dem ~ log(wealth) + polarization + corruption +
    log(women_emp) + I(log(women_emp)^2) + region,
  data = model_data2
)

check_model(democracy_model3)
```

## Your Turn! {.smaller}

<br>

- Run `check_model()` on a model from your own project
- Which LINE assumptions look satisfied? Which look violated?
- Try a log transformation on a skewed predictor — does it improve the diagnostics?

```{r}
#| label: timer3

countdown(minutes = 10, id = "timer3", right = "10%", bottom = "5%")
```
