---
title: Hypothesis Testing 2
subtitle: Associations between Variables
date: today
date-format: long
footer: "[IAFF 6501 Website](https://quant4ia.rocks/)"
logo: images/iaff6501-logo.png
format:
  revealjs:
    theme: [simple, custom.scss]
    transition: fade
    slide-number: true
    #multiplex: true
    chalkboard: true
execute:
  echo: true
  message: false
  warning: false
  freeze: auto
---


## Hypotheses

<br>

- **Null hypothesis:** there is no relationship between treatment and outcome, the difference is due to chance
- **Alternative hypothesis:** there is a relationship, the difference is not due to chance

## Approach

<br>

- Under the null hypothesis, treatment has **NO** impact on *y* (the outcome)
- This means that if we were to change the values of the treatment variable, the values on *y* **would stay the same**

## Approach

- So...we can simulate the null distribution by:
    - Reshuffling the treatment variable
    - Calculating the treatment effect
    - Repeating many times
    
. . . 

- Then we can ask: how likely would we be to observe the treatment effect in our data, *if there is no effect of the treatment*?

## Résumé Experiment Example

<br>

Bertrand and Mullainathan studied racial discrimination in responses to job applications in Chicago and Boston. They sent 4,870 résumés, randomly assigning names associated with different racial groups.

<br>

```{r}
library(openintro)
myDat <- resume 
```

## Callbacks by Race

<br>

Remember, race of applicant is randomly assigned.

```{r}
library(tidyverse)

mns <- myDat |>
  group_by(race) |> 
  summarize(calls = mean(received_callback))
mns
```

##

<br> 

Let's save the means for white and black applicants.

<br>

```{r}
mean_white = mns$calls[2]
mean_black = mns$calls[1]
```

##

<br>

And calculate the treatment effect. The treatment effect is the difference in means.

<br>

```{r}
teffect <- mean_white - mean_black
teffect
```

## {.center}

Before formal tests, let's look at the data--the estimates **and** the confidence intervals...

##

<br>

First, let's make the CIs for the white applicants.

<br>

```{r}
library(tidymodels)
boot_df_white <- myDat |>
  filter(race == "white") |>
  specify(response = received_callback) |>
  generate(reps = 15000, type = "bootstrap") |>
  calculate(stat = "mean")
ci_white <- boot_df_white |> get_ci(level = 0.95)
ci_white
```

##

<br>

Visualize the bootstrap distribution and CI for white applicants.

<br>

```{r}
#| code-fold: true
boot_df_white |>
  visualize() +
  shade_ci(ci_white, color = "red", fill = "red") +
  labs(
    title = "Distribution of the Means of the Bootstrap Samples",
    x = "Mean",
    y = "Count"
  ) +
  theme_minimal()
```

##

<br>

Now, let's create the CIs for black applicants.

<br>

```{r}
boot_df_black <- myDat |>
  filter(race == "black") |>
  specify(response = received_callback) |>
  generate(reps = 15000, type = "bootstrap") |>
  calculate(stat = "mean")
ci_black <- boot_df_black |> get_ci(level = 0.95)
ci_black
```

##

<br>

Visualize the bootstrap distribution and CI for black applicants.

<br>

```{r}
#| code-fold: true
boot_df_black |>
  visualize() +
  shade_ci(ci_black, color = "red", fill = "red") +
  labs(
    title = "Distribution of the Means of the Bootstrap Samples",
    x = "Mean",
    y = "Count"
  ) +
  theme_minimal()
```

## 

<br>

Now, let's tidy the data for plotting.

<br>

```{r}
plotData <- tibble(
  race = c("Black", "White"),
  meanCalls = c(mean_black, mean_white),
  lower95 = c(ci_black$lower_ci, ci_white$lower_ci),
  upper95 = c(ci_black$upper_ci, ci_white$upper_ci)
)
plotData
```

## Plot

```{r}
#| echo: false

ggplot(plotData, aes(y = meanCalls, x = race, ymin = lower95, ymax = upper95)) +
  geom_col(fill = "steelblue4") +
  geom_errorbar(width = .05) +
  theme_bw()  +
 ylim(0, .15) +
  labs(x = "Race of Applicant",
       y = "Call Back Rate")
```

## Plot

<br>

```{r}
#| echo: true
#| eval: false

ggplot(plotData, aes(y = meanCalls, x = race, ymin = lower95, ymax = upper95)) +
  geom_col(fill = "steelblue4") +
  geom_errorbar(width = .05) +
  theme_bw()  +
  ylim(0, .15) +
  labs(x = "Race of Applicant",
       y = "Call Back Rate")
```

## Is this evidence of racial discrimination?

<br>

- What is the null hypothesis?

. . .

- What is the alternative hypothesis?

. . .

- How can we formally test the null hypothesis to decide whether to reject it?

## Formal Hypothesis Test

<br>

::: incremental
- Calculate the difference in means (White - Black)
- Shuffle the race variable
- Calculate the difference in means for the shuffled data
- Repeat many times
- Simulates the null distribution of **differences** in callbacks
:::

## Hypothetical Original Data 

<br>

| Applicant | Race  | Callback |
|-----------|-------|----------|
| A         | Black | Yes      |
| B         | Black | No       |
| C         | Black | No       |
| D         | White | Yes      |
| E         | White | No       |
| F         | White | No       |

## Step 1: Calculate Original Difference in Callback Rates

<br> 

- **Objective:** Understand initial association between race and callback rates

## Step 2: Shuffle (Permute) the Race Variable

<br>

- **Method:** Randomly reassign race labels, keeping callback outcomes fixed

## Hypothetical Shuffled Data

<br>

| Applicant | Race (Shuffled) | Callback |
|-----------|-----------------|----------|
| A         | White           | Yes      |
| B         | Black           | No       |
| C         | White           | No       |
| D         | White           | Yes      |
| E         | Black           | No       |
| F         | Black           | No       |

## Step 3: Calculate Difference in Callback Rates Again

<br> 

- **After Shuffling:** Calculate the difference in callback rates between Black and White groups
- **Purpose:** Determine if observed difference is due to chance

## Repeat Many Times

<br>

- Repeat shuffling 5000 times to generate a distribution of differences by chance
- **Test:** Compare observed difference to null distribution to assess effect of race on callbacks
- If observed difference is extreme (p-value is low), reject the null hypothesis

## Simulating with `tidymodels`

<br>

In real life we are going to use the `tidymodels` package to do the simulation for us.

```{r}
null_dist <- myDat |>
  specify(response = received_callback, explanatory = race) |>
  hypothesize(null = "independence") |>
  generate(5000, type = "permute") |>
  calculate(stat = "diff in means", 
            order = c("white", "black")) # 
```

## Calculate the p-value

<br>

```{r}
null_dist |>
  get_p_value(obs_stat = teffect, direction = "greater")
```

## Visualize the p-value

```{r}
#| echo: true
#| code-fold: true

null_dist |>
  visualize() +
  shade_p_value(obs_stat = teffect, direction = "greater")
```

## What should we conclude?

<br>

::: incremental
- The p-value is very small (below .05 threshold)
- Therefore, we **reject** the null hypothesis: the racial gap is extremely unlikely to have occurred due to chance alone
- This is evidence of racial discrimination
:::

## Your Turn! {.smaller}

<br>

- Use the **gender** variable in the `resume` data to assess whether there is gender discrimination in call backs
- Plot means and 95% confidence intervals for the call back rate for men and women
- Write the null and alternative hypotheses
- Simulate the null distribution
- Visualize the null distribution and the gender gap
- Calculate the p-value
- What do you conclude from your test?  

