---
title: "Lab 4"
subtitle: "Bootstrap Confidence Intervals and a Single-Proportion Test"
author: "YOUR NAME HERE"
format: html
params:
  completed: false
---

::: {.callout-tip title="How to complete this lab"}
Fill in each `???` with the correct code. Once **all** placeholders are filled in, change `completed: false` to `completed: true` in the YAML header above and render to HTML. For your final submission, change `format: html` to `format: pdf`.
:::

## Overview

Use a simulation-based workflow (bootstrap CI + hypothesis test) to analyze a governance reform example.

You are encouraged to have the second half of Lesson 6.2 and Lesson 7.1 open while completing this lab.

## Getting Started

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(tidymodels)
run <- isTRUE(params$completed)
```

::: {.callout-note}
If `tidymodels` is not installed locally, run `install.packages("tidymodels")`.
:::

## The Data

A municipal permit office introduces a transparency reform (posted fee schedules + anonymous reporting) to reduce unofficial payment requests.

In a sample of 50 permit applicants after the reform, 6 reported being asked for an unofficial payment and 44 did not.

```{r}
#| label: permit-data

permit_office <- tibble(
  outcome = c(rep("asked_payment", 6), rep("not_asked", 44))
)
```

## Part 1: Explore the Sample (25 points)

### Sample summary (10 pts)

Write code to:

- view the first 6 rows
- count outcomes
- calculate the observed sample payment-request rate (proportion `asked_payment`)

```{r}
#| label: inspect-summary

# Write your code here

```

**Question:** What is the observed sample payment-request rate (proportion and percent)?

YOUR ANSWER HERE

### Bar chart + short interpretation (15 pts)

Create a bar chart of `outcome`.

- include a title and axis labels
- use `theme_minimal()`

```{r}
#| label: sample-bar-chart

# Write your plot code here

```

**Question:** Based only on this sample, does the reform look promising? Why is this not enough to make a strong causal claim?

YOUR ANSWER HERE

## Part 2: Bootstrap Confidence Interval (45 points)

Estimate a 95% bootstrap confidence interval for the payment-request rate.

### Bootstrap distribution (20 pts)

Fill in the blanks:

```{r}
#| label: bootstrap-dist
#| eval: !expr run

set.seed(66)

boot_df <- permit_office |>
  specify(response = ???, success = ???) |>
  generate(reps = 5000, type = ???) |>
  calculate(stat = ???)

boot_df
```

### CI estimate and plot (15 pts)

Use `get_ci()` and `shade_ci()` to estimate and visualize the 95% CI.

```{r}
#| label: bootstrap-ci-plot
#| eval: !expr run

ci_95 <- boot_df |>
  get_ci(level = 0.95)

boot_df |>
  visualize() +
  ??? +
  labs(
    title = "Bootstrap Distribution of the Payment-Request Rate",
    x = "Sample proportion asked for unofficial payment",
    y = "Count"
  ) +
  theme_minimal()
```

### CI interpretation (10 pts)

Write 2-3 sentences interpreting the 95% confidence interval in context.

YOUR INTERPRETATION HERE

## Part 3: Single-Proportion Hypothesis Test (50 points)

Suppose the office's historical payment-request rate is 20%, and the null hypothesis says the reform had no effect.

### Hypotheses (10 pts)

Question: Is the payment-request rate after the reform **lower** than 20%?

- Null hypothesis ($H_0$):
- Alternative hypothesis ($H_A$):

YOUR ANSWERS HERE

### Simulate the null distribution (20 pts)

Fill in the blanks for a simulation under $H_0: p = 0.20$.

```{r}
#| label: null-distribution
#| eval: !expr run

set.seed(71)

null_dist <- permit_office |>
  specify(response = ???, success = ???) |>
  hypothesize(null = "point", p = c("asked_payment" = ???, "not_asked" = ???)) |>
  generate(reps = 2000, type = ???) |>
  calculate(stat = ???)

null_dist
```

### Null distribution plot (5 pts)

Create a histogram of `null_dist` and add a vertical line at the observed sample proportion.

```{r}
#| label: null-dist-plot

# Write your plot code here

```

### One-sided p-value (10 pts)

Because $H_A$ is "lower than 0.20," calculate the proportion of simulated values less than or equal to the observed sample proportion.

```{r}
#| label: p-value
#| eval: !expr run

null_dist |>
  filter(stat <= ???) |>
  summarize(p_value = ???)
```

### Conclusion (5 pts)

Using $\alpha = 0.05$, state whether you reject or do not reject $H_0$, then explain your conclusion in 2-3 sentences.

YOUR CONCLUSION HERE

## Submission (20 points)

1. Replace "YOUR NAME HERE" with your name
2. Make sure all code chunks run without errors
3. Render to PDF
4. Submit the PDF to Blackboard

## Hints (Short Version)

**Observed proportion**

```
permit_office |>
  summarize(p_hat = mean(outcome == "asked_payment"))
```

**Bootstrap workflow**

```
permit_office |>
  specify(response = outcome, success = "asked_payment") |>
  generate(reps = 5000, type = "bootstrap") |>
  calculate(stat = "prop")
```

**Null distribution workflow**

```
permit_office |>
  specify(response = outcome, success = "asked_payment") |>
  hypothesize(null = "point", p = c("asked_payment" = 0.20, "not_asked" = 0.80)) |>
  generate(reps = 2000, type = "draw") |>
  calculate(stat = "prop")
```

**Simulation p-value**

```
null_dist |>
  filter(stat <= 0.12) |>
  summarize(p_value = n() / nrow(null_dist))
```
