---
title: "Lab 3"
subtitle: "Describing Distributions"
author: "YOUR NAME HERE"
format: html
params:
  completed: false
---

::: {.callout-tip title="How to complete this lab"}
Fill in each `???` with the correct code. Once **all** placeholders are filled in, change `completed: false` to `completed: true` in the YAML header above and render to HTML. For your final submission, change `format: html` to `format: pdf`.
:::

## Overview

In this lab, you will practice describing distributions using the V-Dem dataset. You will:

1. Choose one V-Dem variable you are interested in and create a clean variable name for it
2. Calculate and interpret measures of central tendency (mean and median)
3. Calculate and interpret measures of spread (range, IQR, and standard deviation)
4. Visualize the distribution using a histogram/density plot and a box plot
5. Render your document to PDF and submit

You are encouraged to have Module 6.1 open while completing this lab.

## Getting Started

Load the required packages.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(vdemdata)
run <- isTRUE(params$completed)
```

::: {.callout-note}
## Installing vdemdata locally

If you are working on your own computer and don't have `vdemdata` installed, you can install it from GitHub. First install the `pak` package, then use it to install `vdemdata`:

```r
install.packages("pak")
pak::pak("vdeminstitute/vdemdata")
```
:::

::: {.callout-note}
## Choosing a variable

Pick one V-Dem variable that interests you. You can browse variable definitions in the V-Dem codebook: [V-Dem Codebook](https://v-dem.net/documents/55/codebook.pdf)
:::

## Part 1: Build Your Data Frame (20 points)

Filter to year 2022, recode region as in class, and select one variable of your choice.

In the code below, replace `your_var_name` with a short name you choose and replace `???` with the V-Dem variable code.

```{r}
#| label: data-setup
#| eval: !expr run

vdem2022 <- vdem |>
  filter(year == 2022) |>
  select(
    country = country_name,
    year,
    region = e_regionpol_6C,
    your_var_name = ???
  ) |>
  mutate(
    region = case_match(region,
      1 ~ "Eastern Europe",
      2 ~ "Latin America",
      3 ~ "Middle East",
      4 ~ "Africa",
      5 ~ "The West",
      6 ~ "Asia"
    )
  )
```

**Question:** Which variable did you choose, and what does it measure?

YOUR ANSWER HERE

## Part 2: Measures of Central Tendency (30 points)

Use your chosen variable name in all code below.

### Step 1: Calculate mean and median (15 pts)

```{r}
#| label: center-summary
#| eval: !expr run

vdem2022 |>
  summarize(
    mean_var = ???,
    median_var = ???
  )
```

### Step 2: Visualize center on the distribution (15 pts)

Create either a histogram or a density plot of your chosen variable, and add vertical lines for both mean and median.

```{r}
#| label: center-plot

# Write your plot code here

```

**Question:** Is the mean larger than the median, smaller, or about the same? What does that suggest about skew?

YOUR ANSWER HERE

## Part 3: Measures of Dispersion and Box Plot (50 points)

### Step 1: Calculate spread statistics (25 pts)

Calculate the following for your chosen variable:

- minimum and maximum
- first quartile (Q1) and third quartile (Q3)
- IQR
- standard deviation

```{r}
#| label: spread-summary
#| eval: !expr run

vdem2022 |>
  summarize(
    min_var = ???,
    max_var = ???,
    q1_var = ???,
    q3_var = ???,
    iqr_var = ???,
    sd_var = ???
  )
```

### Step 2: Create a box plot (15 pts)

Create one overall box plot of your chosen variable.

```{r}
#| label: overall-boxplot

# Write a single box plot for your variable here

```

### Step 3: Interpret spread and outliers (10 pts)

In 3-4 sentences, describe what you learned from your spread statistics and box plot.

1. Are values tightly clustered or widely dispersed?
2. Does the box plot suggest potential outliers?
3. How does this compare to what you saw in your histogram/density plot?

YOUR INTERPRETATION HERE

## Render as PDF and Submit Your Work (20 pts)

1. Replace "YOUR NAME HERE" at the top with your actual name
2. Make sure all your code chunks run without errors
3. Click "Render" to create your PDF
4. Submit the PDF to Blackboard

## Hints

*Only look at these if you're stuck!*

**Hint 1 - Filtering to one year:**

```
filter(year == 2022)
```

**Hint 2 - Mean and median (replace `your_var_name`):**

```
summarize(
  mean_var = mean(your_var_name, na.rm = TRUE),
  median_var = median(your_var_name, na.rm = TRUE)
)
```

**Hint 3 - Histogram with reference lines (replace `your_var_name`):**

```
ggplot(vdem2022, aes(x = your_var_name)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = mean(vdem2022$your_var_name, na.rm = TRUE), color = "orange") +
  geom_vline(xintercept = median(vdem2022$your_var_name, na.rm = TRUE), color = "darkgreen", linetype = "dashed") +
  theme_minimal()
```

**Hint 4 - Spread statistics (replace `your_var_name`):**

```
summarize(
  min_var = min(your_var_name, na.rm = TRUE),
  max_var = max(your_var_name, na.rm = TRUE),
  q1_var = quantile(your_var_name, 0.25, na.rm = TRUE),
  q3_var = quantile(your_var_name, 0.75, na.rm = TRUE),
  iqr_var = IQR(your_var_name, na.rm = TRUE),
  sd_var = sd(your_var_name, na.rm = TRUE)
)
```

**Hint 5 - Single box plot (replace `your_var_name`):**

```
ggplot(vdem2022, aes(x = "", y = your_var_name)) +
  geom_boxplot(fill = "steelblue") +
  theme_minimal()
```
