---
title: "Lab 6"
subtitle: "Multiple Regression and Outliers"
author: "YOUR NAME HERE"
format: html
params:
  completed: false
---

::: {.callout-tip title="How to complete this lab"}
Fill in each `???` with the correct code. Once **all** placeholders are filled in, change `completed: false` to `completed: true` in the YAML header above and render to HTML. For your final submission, change `format: html` to `format: pdf`.
:::

## Overview

In this lab you will extend the simple regression skills from Lab 5 to multiple regression, and then practice identifying and handling outliers. You will:

1. Fit a multiple regression model and interpret coefficients while holding other variables constant
2. Include a categorical predictor and interpret dummy variable coefficients
3. Identify influential observations using regression diagnostics
4. Compare model results across different outlier-handling strategies

You are encouraged to have the lecture slides for Weeks 10.1 and 10.2 open while completing this lab.

## Getting Started

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(vdemdata)
library(broom)
run <- isTRUE(params$completed)
```

::: {.callout-note}
## Installing vdemdata locally

If you are working on your own computer and don't have `vdemdata` installed, you can install it from GitHub. First install the `pak` package, then use it to install `vdemdata`:

```r
install.packages("pak")
pak::pak("vdeminstitute/vdemdata")
```
:::

## The Data

You will continue working with V-Dem cross-national data for the year 2019. The research question is: **What predicts a country's level of liberal democracy?**

The key variables are:

- `lib_dem` — Liberal Democracy Index (0–1, higher = more democratic)
- `wealth` — GDP per capita (in thousands of USD)
- `log_wealth` — log-transformed GDP per capita
- `polarization` — Political polarization index (higher = more polarized)
- `region` — World region (6 categories)

## Part 1: Multiple Regression (60 points)

### Step 1: Wrangle the data (10 pts)

Fill in the `???` to filter V-Dem to the year 2019 and select the relevant variables.

```{r}
#| label: wrangle
#| eval: !expr run

model_data <- vdem |>
  filter(year == ???) |>
  select(
    country = country_name,
    lib_dem = v2x_libdem,
    wealth  = e_gdppc,
    polarization = v2cacamps,
    region = e_regionpol_6C
  ) |>
  mutate(
    log_wealth = log(wealth),
    region = factor(region,
      labels = c("Eastern Europe", "Latin America", "MENA",
                 "Sub-Saharan Africa", "Western Europe & N. America",
                 "Asia and Pacific"))
  ) |>
  filter(!is.na(lib_dem), !is.na(wealth))

glimpse(model_data)
```

**Question:** How many countries are in the dataset after filtering out missing values?

YOUR ANSWER HERE

### Step 2: Fit a model with two predictors (20 pts)

Fit a multiple regression model predicting `lib_dem` from `log_wealth` and `polarization`. Store the result as `model1` and view the output with `summary()`.

```{r}
#| label: model1
#| eval: !expr run

model1 <- lm(??? ~ ??? + ???, data = model_data)

summary(model1)
```

### Step 3: Interpret the coefficients (20 pts)

**Question:** Write out the estimated regression equation using the coefficients from the output (round to two decimal places).

$$\widehat{LibDem}_i = a + b_1 \times LogWealth + b_2 \times Polarization$$

YOUR EQUATION HERE

**Question:** Interpret the coefficient on `log_wealth`. What does "holding polarization constant" mean in plain language?

YOUR ANSWER HERE

**Question:** Interpret the coefficient on `polarization`. Is the direction of the relationship what you would expect? Why or why not?

YOUR ANSWER HERE

### Step 4: Add a categorical predictor (10 pts)

Now add `region` as a third predictor. Store this as `model2` and view the summary.

```{r}
#| label: model2
#| eval: !expr run

model2 <- lm(lib_dem ~ log_wealth + polarization + region, data = model_data)

summary(model2)
```

**Question:** What is the baseline (reference) region in this model? How do you know?

YOUR ANSWER HERE

**Question:** Pick one region coefficient and interpret it in a sentence. What does it tell us about democracy in that region compared to the baseline?

YOUR ANSWER HERE

## Part 2: Outliers and Influential Observations (40 points)

### Step 1: Identify influential points (20 pts)

Use `broom::augment()` to compute regression diagnostics for `model1`. Then flag observations that are high-leverage, high-residual, or highly influential.

```{r}
#| label: diagnostics
#| eval: !expr run

model_diagnostics <- augment(model1, data = model_data) |>
  mutate(
    high_leverage  = .hat > 2 * mean(.hat),
    high_residual  = abs(.std.resid) > 2,
    high_influence = .cooksd > 4 / nrow(model_data)
  )

model_diagnostics |>
  filter(high_leverage | high_residual | high_influence) |>
  select(country, wealth, lib_dem, high_leverage, high_residual, high_influence)
```

**Question:** How many countries are flagged as influential? Name two that stand out and explain why they might be outliers in this model.

YOUR ANSWER HERE

### Step 2: Compare models with and without influential points (20 pts)

Remove the influential observations and re-fit the model. Compare the coefficients and R-squared between the two versions.

```{r}
#| label: compare-models
#| eval: !expr run

model_full <- lm(lib_dem ~ log_wealth + polarization, data = model_data)

model_clean <- model_diagnostics |>
  filter(!high_leverage & !high_residual & !high_influence) |>
  lm(lib_dem ~ log_wealth + polarization, data = _)

tibble(
  model     = c("Full model", "Outliers removed"),
  intercept = c(coef(model_full)[1],    coef(model_clean)[1]),
  log_wealth  = c(coef(model_full)[2],  coef(model_clean)[2]),
  polarization = c(coef(model_full)[3], coef(model_clean)[3]),
  r_squared = c(glance(model_full)$r.squared, glance(model_clean)$r.squared)
)
```

**Question:** How much do the coefficients change after removing influential points? Does this suggest the outliers were having a large or small effect on the results? What does this tell you about the robustness of the findings?

YOUR ANSWER HERE

## Render as PDF and Submit Your Work

1. Replace "YOUR NAME HERE" at the top with your actual name
2. Make sure all code chunks run without errors
3. Click "Render" to create your PDF
4. Submit the PDF to Blackboard

---

## Bonus: Checking Model Assumptions (up to 10 points)

*This section is optional and corresponds to the optional Module 10.3 on model assumptions.*

Use the `performance` package to run automated diagnostics on `model2` (the model with region included).

```{r}
#| label: check-model
#| eval: !expr run
#| message: false

library(performance)

check_model(model2)
```

**Question:** Look at the **Linearity** panel. Does the smoother line stay close to zero across fitted values, or does it show a curve? What would a curve indicate?

YOUR ANSWER HERE

**Question:** Look at the **Homogeneity of Variance** panel. Does the spread of residuals look roughly constant, or does it fan out? What would a funnel shape indicate?

YOUR ANSWER HERE

**Question:** Look at the **Normality of Residuals** panel. Do the points follow the diagonal line closely? Are there any systematic departures in the tails?

YOUR ANSWER HERE

---

## Hints

*Only look at these if you're stuck!*

**Hint 1 — Fitting a multiple regression model:**

```r
model1 <- lm(lib_dem ~ log_wealth + polarization, data = model_data)
summary(model1)
```

**Hint 2 — Including a factor variable as a predictor:**

```r
model2 <- lm(lib_dem ~ log_wealth + polarization + region, data = model_data)
```
R automatically creates dummy variables for each level of `region`, leaving the first level as the reference category.

**Hint 3 — Running regression diagnostics:**

```r
model_diagnostics <- augment(model1, data = model_data) |>
  mutate(
    high_leverage  = .hat > 2 * mean(.hat),
    high_residual  = abs(.std.resid) > 2,
    high_influence = .cooksd > 4 / nrow(model_data)
  )
```

**Hint 4 — Removing outliers and re-fitting:**

```r
model_clean <- model_diagnostics |>
  filter(!high_leverage & !high_residual & !high_influence) |>
  lm(lib_dem ~ log_wealth + polarization, data = _)
```

**Hint 5 — Comparing coefficients with `glance()` and `coef()`:**

`coef(model)` extracts the coefficients as a named vector. `glance(model)$r.squared` extracts R-squared from the `broom` package.
