---
title: "Lab 7"
subtitle: "Logistic Regression and Conflict Onset"
author: "YOUR NAME HERE"
format: html
params:
  completed: false
---

::: {.callout-tip title="How to complete this lab"}
Fill in each `???` with the correct code. Once **all** placeholders are filled in, change `completed: false` to `completed: true` in the YAML header above and render to HTML. For your final submission, change `format: html` to `format: pdf`.
:::

## Overview

In this lab you will practice fitting and interpreting logistic regression models using data from the peace science literature. You will:

1. Build a conflict dataset using the `peacesciencer` package and understand its structure
2. Fit a logistic regression model with `glm()`, interpret log-odds and odds ratios, and calculate predicted probabilities

You are encouraged to have the lecture slides for Weeks 11.1 and 11.2 open while completing this lab.

## Getting Started

```{r}
#| label: setup
#| message: false
#| warning: false

library(peacesciencer)
library(marginaleffects)
library(broom)
library(tidyverse)
run <- isTRUE(params$completed)
```

::: {.callout-note}
## Installing peacesciencer

If you do not have `peacesciencer` installed, run the following in your console:

```r
install.packages("peacesciencer")
```

Similarly for `marginaleffects`:

```r
install.packages("marginaleffects")
```
:::

## The Data

You will work with country-year data on civil conflict from the Uppsala Conflict Data Program (UCDP), combined with economic and political variables from several standard datasets. The research question is: **Does national wealth reduce the likelihood of civil conflict onset?**

The key variables are:

- `ucdponset` — Conflict onset (1 = a new civil conflict began that year, 0 = no onset)
- `wbgdppc2011est` — Log GDP per capita (World Bank, 2011 USD)
- `gw_name` — Country name (Gleditsch-Ward system)

## Part 1: Set Up the Data (30 points)

### Step 1: Build the conflict dataset (15 pts)

Fill in the `???` to complete the pipeline. You want country-years from the Gleditsch-Ward state system, filtered to 1946–1999, with UCDP intrastate conflict data and the four add-on variables shown below.

```{r}
#| label: load-data
#| eval: !expr run

conflict_df <- create_stateyears(system = ???) |>
  filter(year %in% c(???:???)) |>
  add_ucdp_acd(type = c("intrastate"), only_wars = FALSE) |>
  add_democracy() |>
  add_creg_fractionalization() |>
  add_sdp_gdp() |>
  add_rugged_terrain()

glimpse(conflict_df)
```

### Step 2: Understand the outcome variable (15 pts)

**Question:** What does each row in `conflict_df` represent? What does a value of `1` in `ucdponset` mean, and why is conflict onset the "success" outcome in a Bernoulli trial framework even though conflict is obviously bad?

YOUR ANSWER HERE

## Part 2: Fit and Interpret a Logistic Regression (70 points)

### Step 1: Fit a bivariate logistic regression (10 pts)

Fill in the `???` to fit a logistic regression predicting `ucdponset` from `wbgdppc2011est`. Use `glm()` with `family = "binomial"`.

```{r}
#| label: bivariate-model
#| eval: !expr run

conflict_model <- glm(??? ~ ???,
                      data = conflict_df,
                      family = ???)

summary(conflict_model)
```

### Step 2: Write out the model equation (15 pts)

Using the `Estimate` column from the output, write out the estimated log-odds equation below (round coefficients to two decimal places):

$$\log\left(\frac{p}{1-p}\right) = a + b \times \text{logGDPpc}$$

YOUR EQUATION HERE

### Step 3: Interpret the coefficient (15 pts)

**Question:** What is the sign of the coefficient on `wbgdppc2011est`? What does this tell us about the relationship between national wealth and conflict onset? Is the effect statistically significant at the 0.05 level?

YOUR ANSWER HERE

### Step 4: Calculate and interpret the odds ratio (15 pts)

Use `tidy()` with `exponentiate = TRUE` to convert the coefficient to an odds ratio.

```{r}
#| label: odds-ratio
#| eval: !expr run

tidy(conflict_model, exponentiate = TRUE) |>
  select(term, estimate, p.value)
```

**Question:** Interpret the odds ratio for `wbgdppc2011est`. For a one-unit increase in log GDP per capita, by how much are the odds of conflict onset multiplied? Does wealthier mean more or less likely to experience conflict onset?

YOUR ANSWER HERE

### Step 5: Calculate predicted probabilities (10 pts)

Use `predictions()` from the `marginaleffects` package to calculate the predicted probability of conflict onset for three countries in the year 1999.

```{r}
#| label: predicted-probs
#| eval: !expr run

selected_countries <- conflict_df |>
  filter(
    gw_name %in% c("United States of America", "Venezuela", "Rwanda"),
    year == 1999
  )

marg_effects <- predictions(conflict_model, newdata = selected_countries)

tidy(marg_effects) |>
  select(estimate, conf.low, conf.high, gw_name)
```

**Question:** What is the predicted probability of conflict onset for each country? Do the differences across countries make intuitive sense given their wealth levels?

YOUR ANSWER HERE

## Render as PDF and Submit Your Work

1. Replace "YOUR NAME HERE" at the top with your actual name
2. Make sure all code chunks run without errors
3. Click "Render" to create your PDF
4. Submit the PDF to Blackboard

---

## Hints

*Only look at these if you're stuck!*

**Hint 1 — Building the conflict dataset:**

```r
conflict_df <- create_stateyears(system = 'gw') |>
  filter(year %in% c(1946:1999)) |>
  add_ucdp_acd(type = c("intrastate"), only_wars = FALSE) |>
  add_democracy() |>
  add_creg_fractionalization() |>
  add_sdp_gdp() |>
  add_rugged_terrain()
```

**Hint 2 — Fitting a logistic regression:**

```r
conflict_model <- glm(ucdponset ~ wbgdppc2011est,
                      data = conflict_df,
                      family = "binomial")
summary(conflict_model)
```

**Hint 3 — Getting odds ratios:**

```r
tidy(conflict_model, exponentiate = TRUE) |>
  select(term, estimate, p.value)
```

An odds ratio below 1 means the predictor is associated with *lower* odds of the outcome; above 1 means *higher* odds.

**Hint 4 — Calculating predicted probabilities:**

```r
selected_countries <- conflict_df |>
  filter(gw_name %in% c("United States of America", "Venezuela", "Rwanda"),
         year == 1999)

tidy(predictions(conflict_model, newdata = selected_countries)) |>
  select(estimate, conf.low, conf.high, gw_name)
```
